Unshuffling Permutations

### 2015-09-25 submission.tex
add nex conclusion

  We now turn to defining various gadgets (sequences of integers)
  that act as building blocks in our construction of a new permutation $\mu$:
  \begin{align*}
  \sigma'  &= ((k+1) \; \sigma \; (k+2)) \; [2N_2 + N_4 + 2n + k + 2] \\
  \pi'     &= ((n+1) \; \pi \; (n+2)) \; [2N_2 + N_4 + n + k + 2] \\
  \sigma'' &= \sigma \; [2N_2 + N_4] \\
  \pi''    &= \pi \; [2N_2 + N_4 + k] \\
  \nu_1    &= \nearrow_{N_1} \; [2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu'_1   &= \nearrow_{N_1} \; [N_1 + 2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu_2    &= \nearrow_{N_2} \; [N_2] \\
  \nu'_2   &= \searrow_{N_2} \\
  \nu_3    &= \nearrow_{N_3} \; [2N_2 + 2N_4 + 2n + 2k + 4] \\
  \nu'_3   &= \nearrow_{N_3} \; [2N_2 + N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu_4    &= \searrow_{N_4} \; [2N_2 + N_4 + 2n + 2k + 4] \\
  \nu'_4   &= \searrow_{N_4} \; [2N_2]\text{.}
  \end{align*}
  We are now in position to define our target permutation $\mu$
  (see Fig.~\ref{fig:reduction} for an illustration):
  $$
  \mu
  =
  \nu_1 \; \nu_2 \; \nu'_1 \; \nu_3 \; \sigma' \; \nu_4 \; \nu'_3 \; \pi' \; \nu'_4 \; \pi'' \; \sigma''
  \text{.}
  $$

@@

There are a number of further directions of investigation in this
general subject. We mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, by
Proposition~\ref{prop:bijection_binary_to_permutations_squares},
how many binary strings of length $2n$ are squares; Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?
Given two permutations $\pi$ and~$\sigma$, how hard is the problem of
deciding whether $\sigma$ is a square root of~$\pi$?

Besides, in this work, we investigate some algebraic properties of the
shuffle of permutations which have direct combinatorial consequences.
However, this work raises several additional classical algebraic
questions. Among these, one can ask for a complete algebraic study of
$\QQ[S]$ as a graded associative algebra for the shuffle  product
$\SHUFFLE$. Describing a generating family for $\QQ[S]$, defining
multiplicative bases of $\QQ[S]$, and determining whether $\QQ[S]$ is
free as an associative algebra are worthwhile questions.

>>>>

  We now turn to defining various gadgets (sequences of integers)
  that act as building blocks in our construction of a new permutation $\mu$:
  \begin{align*}
  \sigma'  &= ((k+1) \; \sigma \; (k+2)) \; [2N_2 + N_4 + 2n + k + 2] \\
  \pi'     &= ((n+1) \; \pi \; (n+2)) \; [2N_2 + N_4 + n + k] \\
  \sigma'' &= \sigma \; [2N_2 + N_4] \\
  \pi''    &= \pi \; [2N_2 + N_4 + k] \\
  \nu_1    &= \nearrow_{N_1} \; [2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu'_1   &= \nearrow_{N_1} \; [N_1 + 2N_2 + 2N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu_2    &= \nearrow_{N_2} \; [N_2] \\
  \nu'_2   &= \searrow_{N_2} \\
  \nu_3    &= \nearrow_{N_3} \; [2N_2 + 2N_4 + 2n + 2k + 4] \\
  \nu'_3   &= \nearrow_{N_3} \; [2N_2 + N_3 + 2N_4 + 2n + 2k + 4] \\
  \nu_4    &= \searrow_{N_4} \; [2N_2 + N_4 + 2n + 2k + 4] \\
  \nu'_4   &= \searrow_{N_4} \; [2N_2]\text{.}
  \end{align*}
  We are now in position to define our target permutation $\mu$
  (see Fig.~\ref{fig:reduction} for an illustration):
  $$
  \mu
  =
  \nu_1 \; \nu_2 \; \nu'_1 \; \nu_3 \; \sigma' \; \nu_4 \; \nu'_3 \; \pi' \; \nu'_4 \; \pi'' \; \sigma''
  \text{.}
  $$

@@

There are a number of further directions of investigation in this
general subject. They cover several areas: algorithmic, combinatorics,
and algebra. Let us mention several -~not necessarily new~- open
problems that are, in our opinion, the most interesting. How many
permutations of $S_{2n}$ are squares? How many $(213,231)$-avoiding
permutations of $S_{2n}$ are squares? (Equivalently, by
Proposition~\ref{prop:bijection_binary_to_permutations_squares},
how many binary strings of length $2n$ are squares; Problem~4
in \cite{Henshall:Rampersad:Shallit:2011})? How hard is the problem of
deciding whether a $(213,231)$-avoiding permutation is a square
(Problem~4 in \cite{Henshall:Rampersad:Shallit:2011},
see also \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013})?
Given two permutations $\pi$ and~$\sigma$, how hard is the problem of
deciding whether $\sigma$ is a square root of~$\pi$?
Moreover, one can ask for a complete algebraic study of
$\QQ[S]$ as a graded associative algebra for the shuffle  product
$\SHUFFLE$. Describing a generating family for $\QQ[S]$, defining
multiplicative bases of $\QQ[S]$, and determining whether $\QQ[S]$ is
free as an associative algebra are worthwhile questions.
